---
title: "lecture4_01_16"
author: "Ye Tao"
date: "1/16/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Today, we will spend some time talking about some special data types in R.
- factors
- data and time


## Factors


When importing data to R, base R has a burning desire to turn character
information into factor. See for example, `read.table`, and `read.csv`.

```{r, message = FALSE}
library(tidyverse)
```

```{r, eval = FALSE}
# to illustrate the issue of `read.csv`, let's write a csv file out of the flights dataset
library(nycflights13)
write_csv(flights %>% sample_n(100), "flights.csv")
```

```{r, eval = FALSE}
# base R function, character variables are automatically converted to factors
read.csv("flights.csv") # all the columns right now are factors when using base r read.csv function

# notice that tailnum are are just character but the r based read.csv functoion will treat it as factor or levels
# which is the thing we do not wanna see
```


```{r, eval = FALSE}
# tidyverse function readr
# readr function, character variables are imported as is
read_csv("flights.csv")


# there are several workarounds,
# 1. we could use `mutate_if` to force the conversion
read_csv("flights.csv") %>%
  mutate_if(is.character, as_factor) # check the content of each column with mutate_if
read_csv("flights.csv") %>%
  mutate_if(~ is.character(.) && n_distinct(.) < 50, as_factor)

# 2. we could specify the column types manually
read_csv("flights.csv", col_types = "iiiddddddfccffddddT") # 19 columns i -> integer, d -> decimal double, f-> factor
# c-> character, T -> time
# 3. use the rstudio import interface just not function to use
```


### Factor inspection

Let's consider the dataset `gss_cat`: A sample of categorical variables from the General Social survey



```{r}
# partyid is factor here different political - preference
class(gss_cat$partyid)
levels(gss_cat$partyid)

# unique value in this factor, not too much difference with level but factor unqiue gives factor in the end 
# levels gives charatcer
fct_unique(gss_cat$partyid)
nlevels(gss_cat$partyid)
```

```{r}
gss_cat %>% count(partyid)
fct_count(gss_cat$partyid, sort = TRUE) #
```



### Combining factors


```{r}
fa <- factor("a")
fb <- factor("b")
fab <- factor(c("a", "b")) # factor for two value

as.integer(fa) # numerical value a represent the level:a

# so when conncant them this will give a sequence if number which is the thing we do not wanna see
c(fa, fb, fab)  # not what you want! 1 1 1 2
fct_c(fa, fb, fab) # combine all the factor, 1this correct
```


### Dropping unused levels


The number of levels won't change even all the rows corresponding to specific factor level are dropped.
wield thing here
```{r}
gss_cat2 <- gss_cat %>%
  filter(partyid %in% c("Independent", "Strong democrat", "Strong republican"))
nlevels(gss_cat2$partyid) # this will keep the level of 10 which is the level of original dataset

```

```{r}
# drop unused levels of a specific factor
gss_cat2$partyid <- gss_cat2$partyid %>% fct_drop() # dropp all other unused data
levels(gss_cat2$partyid)

# or
gss_cat %>% mutate(partyid = fct_drop(partyid))
```

```{r}
# drop unused levels for all the factors in a data frame
gss_cat2 <- gss_cat2 %>% droplevels()
```

### Change order of the levels

```{r}
gss_cat$partyid %>%
  levels()
## order factor by frequency
gss_cat %>% mutate(partyid = partyid %>% fct_infreq())
# this helps to visualize by order in frequency
## backwards!
gss_cat %>% mutate(partyid = partyid %>% fct_infreq() %>% fct_rev()) %>% pull(partyid) %>% levels
```

Why?

```{r}
ggplot(gss_cat) + geom_bar(aes(partyid)) + coord_flip()
```

```{r}

# reverse the order fct_rev
ggplot(gss_cat) + geom_bar(aes(partyid %>% fct_infreq() %>% fct_rev())) +
  xlab("partyid") + coord_flip()
```
https://www.sharpsightlabs.com/blog/highlight-data-in-ggplot2/
hight the one single bar in the bar chart





fct_reorder used in david robinson check out the youtube channel



```{r}
gss_cat %>%
  group_by(partyid) %>%
  summarize(mean_age = mean(age, na.rm = TRUE)) %>%
  ggplot(aes(x = mean_age, y = partyid)) + # reorder factor needed here
  geom_point() + ylab("partyid")
```


```{r}
# reorder factor according to values of another variable
# age increease the strong poitical idea
gss_cat %>%
  group_by(partyid) %>%
  summarize(mean_age = mean(age, na.rm = TRUE)) %>%
  ggplot(aes(x = mean_age, y = fct_reorder(partyid, mean_age))) + # reorder factor  partyid by mean age with increasing order
  geom_point() + ylab("partyid")
```


### Change to any order

```{r}
gss_cat$partyid %>% levels()
gss_cat$partyid %>%
  fct_relevel("Strong republican", "Strong democrat") %>% # move factor position group#1 strong republixan
  levels()
# use mutate verb to modifly the data frame to change the columns
gss_cat %>% mutate(partyid = partyid %>% fct_relevel("Strong republican", "Strong democrat")) # ordering factor as you want
```


### Recode levels

```{r}
gss_cat$partyid %>% levels()
gss_cat$partyid %>%
  fct_recode(
    "Independent,near rep" = "Ind,near rep",
    "Independent,near dem" = "Ind,near dem"
  ) %>%
  levels()
# if we need to modifly the data frame, then
gss_cat %>% mutate(partyid = partyid %>%
  fct_recode(
    "Independent,near rep" = "Ind,near rep",
    "Independent,near dem" = "Ind,near dem"
  ))
```

### Collapse levels

```{r}
# collapse small levels automatically
# fct_lump 5 + 1 categoty which is other category
# https://forcats.tidyverse.org/reference/fct_lump.html
gss_cat %>%
  mutate(partyid = partyid %>% fct_lump(5)) %>%
  ggplot() + geom_bar(aes(partyid))
```

```{r}
# collapse manually
gss_cat %>%
  mutate(partyid = partyid %>% fct_collapse(
    Missing = c("No answer", "Don't know"),
    Rep = c("Strong republican", "Not str republican"),
    Ind = c("Ind,near rep", "Independent", "Ind,near dem"),
    Dem = c("Not str democrat", "Strong democrat"),
    Other = c("Other party")
  )) %>%
  ggplot() + geom_bar(aes(partyid))
```

Remark: there is a bug in forcats v0.4.0 such that the argument `group_other` in `fct_collapse` is malfunction.



## Date and time

```{r}
library(lubridate)
today()
now() # default in UTC (Coordinated Universal Time) # international line
```

```{r}
# internally, time is stored as the so called "unix time"
# the number of seconds since 1970-01-01 00:00:00 UTC
as.double(now())
as_datetime(1579192970)
```

```{r}
ymd("2017-01-31")
mdy("January 31st, 2017")
dmy("31-Jan-2017")
```
```{r}
ymd_hms("2017-01-31 20:11:59")
mdy_hm("01/31/2017 08:01")
mdy_hm("01/31/2017 08:01", tz = "US/Eastern") # time zone we have to specify the time zone other wise it will automatically give UTC time zone; I
```

```{r}
# OlsonNames() prints all the time zones 
# for the us time zones
OlsonNames() %>% keep(~str_starts(., "US/"))
```

```{r}
# origin time - UTC in default
(t1 <- mdy_hm("01/31/2017 08:01", tz = "US/Eastern")) # eastern time
# convert timezone
with_tz(t1, tzone = "US/Pacific")
# fix a timezone into ptc time zone
force_tz(t1, tzone = "US/Pacific")
```

### From individual components
```{r}
library(nycflights13)
flights %>%
  select(year, month, day, hour, minute)
```

```{r}

# notice that time zone is not accurate
(flights_dt <- flights %>%
  select(year, month, day, hour, minute) %>%
  mutate(
    date = make_date(year, month, day),
    time = make_datetime(year, month, day, hour, minute, tz = "US/Eastern") # create variable time  - specify the time zone; it is important
  ))
```

Remark: something was wrong above!

```{r}
flights_dt %>% pull(time) %>% head # now it si EST
```




### Get components

```{r}
datetime <- ymd_hms("2016-07-08 12:34:56")
year(datetime)
month(datetime)
month(datetime, label = TRUE)
mday(datetime) # day of the month
yday(datetime) # day of the year
wday(datetime)
wday(datetime, label = TRUE, abbr = FALSE)
dst(datetime) # is it day light savin gtime?




```


```{r}
flights_dt %>%
  mutate(wday = wday(time, label = TRUE)) %>%
  ggplot(aes(x = wday)) +
  geom_bar()
```


### Math on data and time

```{r}
nor <- ymd_hms("2018-01-01 01:30:00",tz="US/Eastern") 
nor + minutes(90)
nor + dminutes(90)


gap <- ymd_hms("2018-03-11 01:30:00",tz="US/Eastern") #2:00am -> 03:00am
gap + minutes(90)
gap + dminutes(90) # duration daylight saving






# Leap Year
leap <- ymd("2019-03-01")
leap + years(1)
leap + dyears(1)


dyears(1) # number seconds 
365 * 24 * 60 * 60
```


Caculate the time difference
```{r}
# with different time zone
t1 <- mdy_hm("01/21/2020 08:31", tz = "US/Eastern")
t2 <- mdy_hm("01/21/2020 12:01", tz = "US/Pacific")
t2 - t1
```

```{r}
# rounding
t1 <- mdy_hm("01/21/2020 08:31", tz = "US/Eastern")
floor_date(t1, unit = "month") # round down the time
round_date(t1, unit = "month")
ceiling_date(t1, unit = "month")
```


### Interval

define the interval to check if a particular time sit inside the interval.
```{r}
intvl <- interval(ymd("2019-12-01"), ymd("2020-01-21"))
int_start(intvl)
int_end(intvl)
ymd("2019-12-31") %within% intvl
```







# References

https://r4ds.had.co.nz
https://lubridate.tidyverse.org/
https://forcats.tidyverse.org/

Terms
Privacy
Security
Status
Help
Contact GitHub
Pricing
API
Training
Blog
About
