---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
```

```{r}
parts <- readr::read_rds(url("https://assets.datacamp.com/production/repositories/5284/datasets/cb649926d41ce73490a9bb710e1501a273061723/parts.rds", "rb"))

part_categories <- readr::read_rds(url("https://assets.datacamp.com/production/repositories/5284/datasets/30fc459770c89e46cce9cce99752ca95fb1d06fe/part_categories.rds", "rb"))
```




```{r}
parts %>% head
```

```{r}
part_categories %>% head
```

```{r}
parts %>%
    inner_join(part_categories, by = c("part_cat_id" = "id"), suffix = c("_parts", "_part_categories")) %>%
    head
```

```{r}
load("flights.Rda")
```






### Six variations on ranking functions

- `row_number`: equivalent to `rank(ties.method = "first")`
- `min_rank`: equivalent to `rank(ties.method = "min")`
- `dense_rank`: like `min_rank()`, but with no gaps between ranks
- `percent_rank`: a number between 0 and 1 computed by rescaling min_rank to [0, 1]
- `cume_dist`: a cumulative distribution function. Proportion of all values less than or equal to the current rank.
- `ntile`: a rough rank, which breaks the input vector into n buckets

```{r}
some_data <- tibble(
  x = c(3, 4, 1, 3, 1)
)
some_data %>% mutate(row_number(), row_number(x), min_rank(x), percent_rank(x))
```


x %>% f() is equivalent to f(x)
x %>% f(.) is also equivalent to f(x)
x %>% f(y) is equivalent to f(x, y)
x %>% f(., y) is equivalent to f(x, y)


### `lead` and `lag`

```{r}
some_data2 <- tibble(
  time = 1:5,
  value = c(3, 4, 1, 3, 1)
)
some_data2 %>% mutate(lag1 = lag(value), lead1 = lead(value))
some_data2 %>% mutate(diff1 = value - lag(value), diff2 = lead(value) - value) %>%
    slice(-c(1:2)) # choose row number
    # slice(n()) # choose the last row
```

### Conditional mutation

```{r}
flights %>% transmute(arr_delay, # only return the new column, mutate gives all the column
  status = if_else(arr_delay > 0, "delayed", "on time")
)
flight_distances <- flights %>%
  transmute(distance,
    distance_type = case_when(  # order matters here, conditional mutate
      distance < 500 ~ "short", # if
      distance < 1000 ~ "mid",  # elif
      TRUE ~ "long"             # else
    )
  )
flight_distances
```

### `recode` values

```{r}
# another way for conditional mutate
flight_distances %>% mutate(distance_type = recode(distance_type,
  long = "long-distance",
  mid = "mid-distance",
  short = "short-distance"
))
```


```{r}
some_data %>% mutate(y = recode(x,
    "1" = "one",
    "3" = "three",
    "4" = "four")
)
```





## `arrange`: changes the ordering of the rows

```{r}
flights %>% arrange(year, month, day)
flights %>% arrange(desc(dep_delay))
```

## `summarize` and `group_by` operations


```{r}
flights %>%
  group_by(tailnum) %>% # tail number
  summarize() # just keep the group but not calculate anthing

flights %>%
  group_by(tailnum) %>% # tail number
  summarize(mean_airtime = mean(air_time, na.rm = TRUE))


# These threee are the same, mapping, reducing into same dataframe
flights %>%
    group_by(tailnum) %>%
    summarise(n = n())

flights %>%
  group_by(tailnum) %>%
  tally() # shorthand

flights %>% count(tailnum) # another shorthand
```

```{r}
flights %>%
  group_by(tailnum) %>%
  summarize(
    count = n(),
    dist = mean(distance, na.rm = TRUE),
    delay = mean(arr_delay, na.rm = TRUE)
  )

flights %>%
  group_by(dest) %>%
  summarize(
    planes = n_distinct(tailnum),
    flights = n()
  ) # same plane goes to mutliple destinations

# group multiple variables
(per_day <- flights %>%
  group_by(year, month, day) %>%
  summarize(flights = n()))

groups(per_day) # summarize used up the day group use groups to check


(per_month <- per_day %>%
  summarize(flights = sum(flights))) # this one used the month group


(per_year <- per_month %>%
  summarize(flights = sum(flights)))
```

## Other useful functions

```{r}
flights %>% glimpse()
flights %>% sample_n(5)
## rows with smallest values of air_time with the original order preserved
flights %>% top_n(3, air_time) # return the top 3 value in air_time
# it returns 4 since there are duplicated value in arrive time
# the dataframe returned doesn't change the orginial order
```

## Some variations of verbs

`tidyverse` also ships with some variations of verbs which could be useful.

```{r}
# only select columns which are numerical
flights %>% select_if(is.numeric) # check the column is numerical not not
flights %>% select_if(~ is.numeric(.)) # ~ means formula, . means each columns here while in the previous example . means flights
# why you wanna have formular here
flights %>% select_if(~ !is.numeric(.))
flights %>% rename_if(is.numeric, toupper) # touper could be some complicated function

flights %>% mutate_if(
  ~ is.numeric(.) && is.double(.), # && and
  round
)

# https://suzan.rbind.io/2018/02/dplyr-tutorial-3/#basic-row-filters
flights %>% filter_if(
  ~ is.numeric(.),
  any_vars(. != 2013) # . means that column
)

# useful for large data set
flights %>% filter_if( # filter the row
  ~ is.numeric(.),
  all_vars(. > 0) # . means that column
)
```

```{r}
flights %>% rename_at(# rename column
  vars(starts_with("arr_")),
  ~ str_replace(., "arr_", "arrival_")
)

# filter_if and filter_at ??
# filter_at medium note: https://medium.com/@kanaugust/you-can-use-filter-if-or-filter-at-5117effd8b3b
flights %>% filter_at( # choose the columns
  vars(ends_with("_time"), -air_time), #=> this the name of column
  all_vars(. >= 1200)
)
```

## Two-table verbs

There are 4 types of joins.

- `inner_join(x, y)` only includes observations that match in both x and y
- `left_join(x, y)` includes all observations in x, regardless of whether they match or not.
- `right_join(x, y)` equivalent to left_join(y, x)
- `full_join(x, y)` includes all observations from x and y

```{r}
df1 <- tibble(id = c(1, 2), v = 2:1)
df2 <- tibble(id = c(1, 3), a = 10, b = "a")
```

```{r}
df1 %>% inner_join(df2)
df1 %>% left_join(df2)
df1 %>% right_join(df2)
df1 %>% full_join(df2)
```


## Tidy Data

There are three interrelated rules which make a dataset tidy

- Each variable must have its own column (long format).
- Each observation must have its own row (wide format).
- Each value must have its own cell.

Using the datasets from R for Data Science to show that the same data could be organised in different ways.



```{r}
# make sure you have tidyr 1.0
library(tidyr)
```

### Privot longer

```{r}
# we need income column since rightnow the data are all the coumn name
relig_income %>%
  pivot_longer(-religion, names_to = "income", values_to = "count")
# no religion but all column, name_to new column name,
```


```{r}
# we need lomg format dataset for dataset analysis
billboard %>%
  pivot_longer( # stack all the columns together making dataset longer for most of time
    cols = starts_with("wk"),
    names_to = "week",
    names_prefix = "wk", # remove the prefix
    names_ptypes = list(week = integer()),
    values_to = "rank",
    values_drop_na = TRUE,
  )
```

### Privot wider

```{r}
fish_encounters %>% pivot_wider( # make dataset wider, spread the dataset for readability for most of time
  names_from = station,
  values_from = seen,
  values_fill = list(seen = 0) # fill missing valu with 0
)
```

```{r}
us_rent_income %>% #spread variable into different value
  pivot_wider(names_from = variable, values_from = c(estimate, moe))
```




## References

- Documentation of dplyr https://dplyr.tidyverse.org/

- R for Data Science http://r4ds.had.co.nz/tidy-data.html











